# Newton's root-finding method

<!-- paragraph -->

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch15-figures/")
```

<!-- paragraph -->

Roots of a function `f(x)` are values `x` such that `f(x)=0`.
<!-- comment -->
Some functions `f`have an explicit expression for their roots.
<!-- comment -->
For example:

<!-- paragraph -->

- the linear function `f(x)=b*x+c=0` has a single root `x=-c/b`, if b is not zero.
<!-- comment -->
- the quadratic function `f(x)=a*x^2+b*x+c=0` has two roots `x=(-b±sqrt(b^2-4*a*c))/(2*a)`, if the discriminant is positive `b^2-4*a*c>0`.
<!-- comment -->
- the sin function `f(x)=sin(a*x)=0` has an infinite number of roots: `pi*z/a` for all integers `z`.

<!-- paragraph -->

However, there are some functions which have no explicit expression for their roots.
<!-- comment -->
For example, the roots of the Poisson loss `f(x)=a*x+b*log(x)+c` have no explicit expression in terms of common mathematical functions.
<!-- comment -->
(actually it has a solution in terms of the [Lambert W function](https://www.wolframalpha.com/input/?i=a*x+%2Bb*log%28x%29%2B+c%3D0) but that function is not commonly available) This goal of this chapter is to create an interactive data visualization that explains [Newton's method](https://en.wikipedia.org/wiki/Newton%27s_method) for finding the roots of such functions.

<!-- paragraph -->

Chapter outline:

<!-- paragraph -->

- We begin by implementing the Newton method for the Poisson loss, to find the root which is larger than the minimum.
<!-- comment -->
We create several static and one interactive data visualization.
<!-- comment -->
- We then suggest an exercise for finding the root which is smaller than the minimum.

<!-- paragraph -->

## Larger root in mean space {#larger-root}

<!-- paragraph -->

We begin by defining coefficients of a Poisson Loss function with two roots.

<!-- paragraph -->

```{r}
Linear <- 95
Log <- -1097
Constant <- 1000
loss.fun <- function(Mean){
  Linear*Mean + Log*log(Mean) + Constant
}
(mean.at.optimum <- -Log/Linear)
(loss.at.optimum <- loss.fun(mean.at.optimum))
library(data.table)
loss.dt <- data.table(mean=seq(0, 100, l=400))
loss.dt[, loss := loss.fun(mean)]
opt.dt <- data.table(
  mean=mean.at.optimum,
  loss=loss.at.optimum,
  point="minimum")
library(animint2)
gg.loss <- ggplot()+
  geom_point(aes(mean, loss, color=point), data=opt.dt)+
  geom_line(aes(mean, loss), data=loss.dt)
print(gg.loss)
```

<!-- paragraph -->

Our goal is to find the two roots of this function.
<!-- comment -->
Newton's root finding method starts from an arbitrary candidate root, and then repeatedly uses linear approximations to find more accurate candidate roots.
<!-- comment -->
To compute the linear approximation, we need the derivative:

<!-- paragraph -->

```{r}
loss.deriv <- function(Mean){
  Linear + Log/Mean
}
```

<!-- paragraph -->

We begin the root finding at a point larger than the minimum,

<!-- paragraph -->

```{r}
possible.root <- mean.at.optimum+1
gg.loss+
  geom_point(aes(
    mean, loss, color=point),
    data=data.table(
      point="start",
      mean=possible.root,
      loss=loss.fun(possible.root)))
```

<!-- paragraph -->

We then use the following implementation of Newton's method to find a root,

<!-- paragraph -->

```{r}
iteration <- 1
solution.list <- list()
thresh.dt <- data.table(thresh=1e-6)
while(thresh.dt$thresh < abs({
  fun.value <- loss.fun(possible.root)
})){
  cat(sprintf("mean=%e loss=%e\n", possible.root, fun.value))
  deriv.value <- loss.deriv(possible.root)
  new.root <- possible.root - fun.value/deriv.value
  solution.list[[iteration]] <- data.table(
    iteration, possible.root, fun.value, deriv.value, new.root)
  iteration <- iteration+1
  possible.root <- new.root
}
root.dt <- data.table(point="root", possible.root, fun.value)
gg.loss+
  geom_point(aes(possible.root, fun.value, color=point),
             data=root.dt)
```

<!-- paragraph -->

The plot above shows the root that was found.
<!-- comment -->
The stopping criterion was an absolute cost value less than `1e-6` so we know that this root is at least that accurate.
<!-- comment -->
The following plot shows the accuracy of the root as a function of the number of iterations.

<!-- paragraph -->

```{r}
solution <- do.call(rbind, solution.list)
solution$new.value <- c(solution$fun.value[-1], fun.value)
gg.it <- ggplot()+
  geom_point(aes(
    iteration, fun.value, color=fun),
    data=data.table(solution, y="fun.value", fun="function"))+
  geom_point(aes(
    iteration, log10(abs(fun.value)), color=fun),
    data=data.table(
      solution, y="log10(abs(fun.value))", fun="function"))+
  scale_color_manual(values=c("function"="black", approximation="red"))+
  geom_point(aes(
    iteration, new.value, color=fun),
    data=data.table(solution, y="fun.value", fun="approximation"))+
  geom_point(aes(
    iteration, log10(abs(new.value)), color=fun),
    data=data.table(
      solution, y="log10(abs(fun.value))", fun="approximation"))+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))+
  facet_grid(y ~ ., scales="free")+
  ylab("")
print(gg.it)
```

<!-- paragraph -->

The plot above shows a horizontal line for the stopping criterion threshold, on the log scale.
<!-- comment -->
It is clear that the red dot in the last iteration is much below that threshold.

<!-- paragraph -->

The plot below shows each step of the algorithm.
<!-- comment -->
The left panels show the linear approximation at the candidate root, along with the root of the linear approximation.
<!-- comment -->
The right panels show the root of the linear approximation, along with the corresponding function value (the new candidate root).

<!-- paragraph -->

```{r}
## y - fun.value = deriv.value * (x - possible.root)
## y = deriv.value*x + fun.value-possible.root*deriv.value
ggplot()+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))+
  facet_grid(iteration ~ step)+
  scale_color_manual(values=c("function"="black", approximation="red"))+
  geom_abline(aes(
    slope=deriv.value, intercept=fun.value-possible.root*deriv.value,
    color=fun),
    data=data.table(solution, fun="approximation", step=1))+
  geom_point(aes(
    new.root, 0, color=fun),
    data=data.table(solution, fun="approximation"))+
  geom_point(aes(
    new.root, new.value, color=fun),
    data=data.table(solution, fun="function", step=2))+
  geom_vline(aes(
    xintercept=new.root, color=fun),
    data=data.table(solution, fun="approximation", step=2))+
  geom_point(aes(
    possible.root, fun.value, color=fun),
    data=data.table(solution, fun="function", step=1))+
  geom_line(aes(
    mean, loss, color=fun),
    data=data.table(loss.dt, fun="function"))+
  ylab("")
```

<!-- paragraph -->

It is clear that the algorithm quickly converges to the root.
<!-- comment -->
The following is an animated interactive version of the same data viz.

<!-- paragraph -->

```{r Ch15-viz}
animint(
  time=list(variable="iteration", ms=2000),
  iterations=gg.it+
    theme_animint(width=300, colspan=1)+
    geom_tallrect(aes(
      xmin=iteration-0.5,
      xmax=iteration+0.5),
      clickSelects="iteration",
      alpha=0.5,
      data=solution),
  loss=ggplot()+
    theme_bw()+
    scale_color_manual(values=c(
      "function"="black",
      approximation="red"))+
    geom_abline(aes(
      slope=deriv.value, intercept=fun.value-possible.root*deriv.value,
      color=fun),
      showSelected="iteration",
      data=data.table(solution, fun="approximation"))+
    geom_point(aes(
      new.root, 0, color=fun),
      showSelected="iteration",
      size=4,
      data=data.table(solution, fun="approximation"))+
    geom_point(aes(
      new.root, new.value, color=fun),
      showSelected="iteration",
      data=data.table(solution, fun="function"))+
    geom_vline(aes(
      xintercept=new.root, color=fun),
      showSelected="iteration",
      data=data.table(solution, fun="approximation"))+
    geom_point(aes(
      possible.root, fun.value, color=fun),
      showSelected="iteration",
      data=data.table(solution, fun="function"))+
    geom_line(aes(
      mean, loss, color=fun),
      data=data.table(loss.dt, fun="function"))+
    ylab(""))
```

<!-- paragraph -->

## Comparison with Lambert W solution {#lambert}

<!-- paragraph -->

The code below uses the Lambert W function to compute a root, and compares its solution to the one we computed using Newton's method.

<!-- paragraph -->

```{r compare-lambert}
inside <- Linear*exp(-Constant/Log)/Log
root.vec <- Log/Linear*c(
  LambertW::W(inside, 0),
  LambertW::W(inside, -1))
loss.fun(c(
  Newton=root.dt$possible.root,
  Lambert=root.vec[2]))
```

<!-- paragraph -->

For these data, the Lambert W function yields a root which is slightly more accurate than our implementation of Newton's method.

<!-- paragraph -->

## Root-finding in the log space {#log-space}

<!-- paragraph -->

The previous section showed an algorithm for finding the root which is larger than the minimum.
<!-- comment -->
In this section we explore an algorithm for finding the other root (smaller than the minimum).
<!-- comment -->
Note that the Poisson loss is highly non-linear as mean goes to zero, so the linear approximation in the Newton root finding will not work very well.
<!-- comment -->
Instead, we equivalently perform root finding in the log space:

<!-- paragraph -->

```{r logloss}
log.loss.fun <- function(log.mean){
  Linear*exp(log.mean) + Log*log.mean + Constant
}
log.loss.dt <- data.table(
  log.mean=seq(-1, 3, l=400)
)[
, loss := log.loss.fun(log.mean)]
ggplot()+
  geom_line(aes(
    log.mean, loss),
    data=log.loss.dt)+
  geom_point(aes(
    log(mean), loss, color=point),
    data=opt.dt)
```

<!-- paragraph -->

**Exercise:**
derive and implement the Newton method for this function, in order to find the root that is smaller than the minimum.
<!-- comment -->
Create an animint similar to the previous section.

<!-- paragraph -->

## Chapter summary and exercises {#Ch15-exercises}

<!-- paragraph -->

In this chapter we explored several visualizations of the Newton method for finding roots of smooth functions.

<!-- paragraph -->

Exercises:

<!-- paragraph -->

- Add a title to each plot.
<!-- comment -->
- Add a size legend to the first plot (black points larger than red), so that we can see when red and black have the same value.
<!-- comment -->
- Add a `geom_hline` to emphasize the loss=0 value in the second plot.
<!-- comment -->
- Add a `geom_hline` to emphasize the stopping threshold in the first plot.
<!-- comment -->
- Turn off one of the two legends, to save space.
<!-- comment -->
- How to specify smooth transitions between iterations?
<!-- comment -->
- Instead of using iteration as the animation/time variable, create a new one in order to show two distinct states/steps for each iteration, i.e.
<!-- comment -->
the `step` variable in the facetted plot above.
<!-- comment -->
- What happens to the rate of convergence when you try to find the larger root in the log space, or the smaller root in the original space?
<!-- comment -->
Theoretically it should not converge as fast, since the functions are more nonlinear for those roots.
<!-- comment -->
Make a data visualization that allows you to select the starting value, and shows how many iterations it takes to converge to within the threshold.
<!-- comment -->
- Create another plot that allows you to select the threshold.
<!-- comment -->
Plot the number of iterations as a function of threshold.
<!-- comment -->
- Derive the loss function for [Binomial regression](https://en.wikipedia.org/wiki/Binomial_regression), and visualize the corresponding Newton root finding method.
<!-- comment -->
- Refactor `gg.it` code to use only one `geom_point`, instead of the four geoms in the current code.
<!-- comment -->
Hint: use `rbind()` to create a single table with all of the data.

<!-- paragraph -->

Next, [Chapter 16](../Ch16/Ch16-change-point.html) explains how to visualize change-point detection models.

<!-- paragraph -->


