# Contents and motivation

<!-- paragraph -->

This manual explains how to design and create interactive data visualizations using the R package `animint2`.

<!-- paragraph -->

## Contents

<!-- paragraph -->

The chapters of this manual are organized as follows.

<!-- paragraph -->

### The `animint2` extensions to the grammar of graphics {#animint}

<!-- paragraph -->

The first seven chapters should be read sequentially, since they give a step by step guide to interactive data visualization using `animint2`.

<!-- paragraph -->

Starting in the [Motivation section](#motivation) below, this first chapter gives an overview of data analysis and visualization.
<!-- comment -->
It provides motivation and a theoretical foundation for the other chapters, and should be especially useful for readers who are completely new to data analysis.
<!-- comment -->
It introduces the method of data visualization prototyping using sketches, without introducing R code.

<!-- paragraph -->

Starting with [chapter 2](ch02), we will show how plot sketches can be translated into R code.
<!-- comment -->
Chapter 2 explains the basics of plotting using ggplots and `animint2`, and should be most useful for readers who have never used `ggplot2`.
<!-- comment -->
It explains how standard ggplots can be rendered on web pages using `animint2`.

<!-- paragraph -->

[Chapter 3](ch03) introduces `showSelected`, one of the two main keywords that `animint2` introduces for interactive data visualization design.
<!-- comment -->
Chapter 3 begins by explaining selection variables, which provide the mechanism of interaction in `animint2`.
<!-- comment -->
Chapter 3 then explains how the`showSelected` keyword makes it possible to plot data subsets.
<!-- comment -->
Chapter 3 also explains how to use smooth transitions and animation.

<!-- paragraph -->

[Chapter 4](ch04) introduces `clickSelects`, the other main keyword that `animint2` introduces for interactive data visualization design.
<!-- comment -->
The`clickSelects` keyword makes it possible for the user to change a selection variable by directly clicking on a plot element.

<!-- paragraph -->

[Chapter 5](ch05) explains several different ways to share your interactive data visualizations on the web.

<!-- paragraph -->

[Chapter 6](ch06) covers some other features of `animint2`, including how to specify hyperlinks, tooltips, data-driven selector variable names.

<!-- paragraph -->

[Chapter 7](ch07) covers the limitations of the current implementation of the `animint2` R package, and explains workarounds for some common issues.
<!-- comment -->
It also includes some ideas for improvements, for those who would like to contribute to `animint2`.

<!-- paragraph -->

### Examples {#examples}

<!-- paragraph -->

The remaining chapters can be read in any order, since each chapter explains how to make data visualizations for a particular data set or statistical algorithm.

<!-- paragraph -->

[Chapter 8](ch08) explains how to create a multi-panel interactive World Bank data visualization.

<!-- paragraph -->

[Chapter 9](ch09) shows a visualization of data from cyclists in Montreal.

<!-- paragraph -->

[Chapter 10](ch10) explains how to create an interactive re-design of the nearest neighbors data visualization from the Elements of Statistical Learning book [@Hastie2009].

<!-- paragraph -->

[Chapter 11](ch11) shows a data visualization that explains the Lasso, a machine learning model for regularized regression.

<!-- paragraph -->

[Chapter 12](ch12) shows a data visualization that explains support vector machines (SVM), a machine learning model for binary classification.

<!-- paragraph -->

[Chapter 13](ch13) explains how to create an interactive visualization that explains the Poisson regression model.

<!-- paragraph -->

[Chapter 14](ch14) shows an example of how to create data-driven selectors using named `clickSelects` and `showSelected` in an interactive visualization of a peak detection model.

<!-- paragraph -->

[Chapter 15](ch15) explains how to create an interactive visualization of the Newton root-finding algorithm.

<!-- paragraph -->

[Chapter 16](ch16) explains how to create an interactive visualization of an optimal changepoint detection model.

<!-- paragraph -->

[Chapter 17](ch17) explains how to create an interactive visualization of the k-means clustering algorithm.

<!-- paragraph -->

[Chapter 18](ch18) explains how to create an interactive visualization of the gradient descent algorithm for learning neural network weight matrices.

<!-- paragraph -->

[Chapter 19](ch19) explains how to visualize P-values, including how to work around overplotting, using a heat map linked to a zoomed scatter plot.

<!-- paragraph -->

### Appendices {#appendices}

<!-- paragraph -->

[Useful idioms](ch99) contains detailed explanations of several R code idioms that are used throughout this manual.

<!-- paragraph -->

[The contributing guide](contribute.html) contains instructions about how you can contribute improvments to this manual.

<!-- paragraph -->

## Motivation {#motivation}

<!-- paragraph -->

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="ch01-figures/")
```

<!-- paragraph -->

The purpose of this manual is to explain the usage of `animint2`, an R package for interactive data visualization.
<!-- comment -->
This introductory chapter answers the following questions:

<!-- paragraph -->

- What are data, and how are they analyzed?
<!-- comment -->
- What is data visualization, and when is it useful for data analysis?
<!-- comment -->
- What is interactive data visualization, and when is it useful?

<!-- paragraph -->

This introductory chapter uses the following outline:

<!-- paragraph -->

- What is data?
<!-- comment -->
- Small data -- data visualization is not necessary.
<!-- comment -->
- Medium data -- static data visualization is sufficient.
<!-- comment -->
- Large data -- interactive data visualization is useful.

<!-- paragraph -->

### What is data analysis?
<!-- comment -->
{#data}

<!-- paragraph -->

Data are any pieces of information that are systematically recorded, either on paper or on a computer.
<!-- comment -->
Anybody can create data, just by systematically writing things down.
<!-- comment -->
Typically, data are created in order to help answer a specific question, and are organized into tables with rows for observations and columns for variables or different types of information.
<!-- comment -->
The word "data" comes from Latin, its plural form "datum," refers to one observation/row of a data table.
<!-- comment -->
We use the term "data set" to refer to a subset of observations/rows, or the entire data table.

<!-- paragraph -->

There are many examples of data that could be created to answer questions based on everyday experiences:

<!-- paragraph -->

- How does the weather this year compare to previous years?
<!-- comment -->
Have we had more or less rain than usual?
<!-- comment -->
To answer these questions, we could create a data table with column for measurements of different weather conditions: temperature, rainfall, etc.
<!-- comment -->
There should also be columns for the date and time of each observation, and a row for each observation.
<!-- comment -->
- How is this new diet affecting me?
<!-- comment -->
If you are trying a new diet, you may want to record what you eat and how you feel after each meal.
<!-- comment -->
In that case you could make a table with a row for each meal and four columns: date, time, what you ate, and how you felt after.
<!-- comment -->
- Does this new lung cancer treatment work better than the old treatment?
<!-- comment -->
A doctor who conducts the clinical trial would randomly assign patients to receive either the new or old treatment.
<!-- comment -->
The doctor would then create a data table with a row for each patient, and several columns: years the patient has smoked, treatment type (new or old), patient age at treatment, patient age at death, tumoral marker measurement, etc.

<!-- paragraph -->

We define "data analysis" as the process of answering these questions by converting the raw data table into other, more comprehensible forms.
<!-- comment -->
One highly effective class of methods for data analysis is called "data visualization," which seeks to provide answers to questions by converting a data set into an informative picture.
<!-- comment -->
The term "data visualization" refers to both the picture itself (also known as a plot, chart, figure, graph, graphic, or data viz), and the process of creating the picture.

<!-- paragraph -->

There are many different ways to perform data analysis, and data sets of different sizes should be analyzed using different techniques.
<!-- comment -->
There are many different ways to characterize the size of data sets, and every author uses a slightly different definition.
<!-- comment -->
In this manual we will use a classification of data sets into three sizes:
small, medium, and large.
<!-- comment -->
We begin by discussing small data sets, for which data visualization is not necessary.

<!-- paragraph -->

### Small data analysis without visualization {#small-data}

<!-- paragraph -->

In this section, we will discuss "small data," which are small enough such that data analysis can be done by simply looking at the entire data table.
<!-- comment -->
For small data sets, there is no need to use data visualization.
<!-- comment -->
Instead, the data can simply be presented for visual inspection in a table.

<!-- paragraph -->

As a concrete example, consider [the famous tea tasting experiment proposed by Ronald Fisher](https://en.wikipedia.org/wiki/Lady_tasting_tea).
<!-- comment -->
A Lady claimed that she could taste the difference when milk is added to the teacup before or after the tea.
<!-- comment -->
Fisher asked the question, can the Lady really taste the difference between the two types of tea?

<!-- paragraph -->

To answer that question, Fisher prepared four cups of tea with milk added after, and four cups of tea with milk added before.
<!-- comment -->
Fisher then placed the cups in a random order, had the Lady taste all eight cups of tea, and asked her to identify the four in which milk was added after the tea.
<!-- comment -->
According to`help(fisher.test)` in R, the Lady correctly identified three of the four cups in which milk was added after the tea.
<!-- comment -->
Fisher than wrote down the following data: the total number of cups (8), the total number of cups with milk added (4), and the total number of cups that the Lady correctly identified (3).
<!-- comment -->
In R, this data set can be viewed by printing a contingency table of count data:

<!-- paragraph -->

```{r, echo=FALSE}
TeaTasting <- matrix(c(
  3, 1, 1, 3),
  nrow = 2,
  dimnames = list(
    Guess = c("Milk", "Tea"),
    Truth = c("Milk", "Tea")))
print(TeaTasting)
```

<!-- paragraph -->

In this case, the data set is small enough such that Fisher's question can be answered by simply looking at the data table itself.
<!-- comment -->
If the Lady had been able to correctly identify all four cups, then that would have been a very convincing demonstration of her ability.
<!-- comment -->
However, she was apparently only able to correctly identify three out of the four cups, which is less convincing.

<!-- paragraph -->

The main topic of this manual is data visualization, which is not necessary for such small data sets.
<!-- comment -->
Instead, we will focus on data sets that are too big to be analyzed by manual visual inspection of the data table.

<!-- paragraph -->

### Medium data analysis with static data visualizations {#medium-data}

<!-- paragraph -->

For medium sized data sets, simply inspecting the data table is no longer sufficient to answer the questions posed during data analysis.
<!-- comment -->
Medium data are big enough such that we need to use visualization to understand the data.

<!-- paragraph -->

For example, consider the following data on atmospheric carbon dioxide (CO2) concentrations, recorded monthly between 1959 and 1997.

<!-- paragraph -->

```{r, echo=FALSE}
year.int <- as.integer(floor(time(co2)))
month.int <- as.integer(time(co2) %% 1 * 12 + 1)
year.month.chr <- sprintf("%d-%02d-15", year.int, month.int)
year.month.POSIXct <- as.POSIXct(strptime(year.month.chr, "%Y-%m-%d"))
hidden <- Sys.setlocale("LC_TIME", "C")
month.str <- strftime(year.month.POSIXct, "%B")
month <- factor(month.str, month.str[1:12])
library(data.table)
co2.details <- data.table(
  year.int,
  month.int,
  month,
  year.month.POSIXct,
  ppm=as.numeric(co2))
co2.details
```

<!-- paragraph -->

Printing these data on the R command line shows that there are 468 rows/observations total.
<!-- comment -->
This is not a huge number of observations, but it is already big enough so that answering questions is not easy by simple visual inspection of the data table.
<!-- comment -->
Instead, we will create a static data visualization:

<!-- paragraph -->

```{r, echo=FALSE}
library(animint2)
extremes <- co2.details[month.int %in% c(5, 10),]
ggplot()+
  geom_line(aes(
    year.month.POSIXct, ppm),
    data=co2.details)+
  geom_point(aes(
    year.month.POSIXct, ppm, color=month),
    data=extremes)+
  xlab("Time of measurement")+
  ylab("Atmospheric carbon dioxide (parts per million)")
```

<!-- paragraph -->

The static data visualization shows that CO2 concentrations increased over the second half of the twentieth century.
<!-- comment -->
This particular data visualization is called a [Keeling Curve](https://en.wikipedia.org/wiki/Keeling_Curve).
<!-- comment -->
It is named after Charles David Keeling, the pioneering scientist who collected the first frequent regular data on atmospheric CO2.
<!-- comment -->
The general increasing trend can be explained by considering the chemical process of combustion, which converts oxygen to CO2.
<!-- comment -->
Keeling noted that "the observed rate of increase is nearly that to be expected from the combustion of fossil fuel" [@keeling1960concentration].

<!-- paragraph -->

The data visualization also reveals the interesting seasonal trend that attains a local maximum each May and a local minimum each October.
<!-- comment -->
This seasonal trend can be explained by considering the forests in the Northern Hemisphere.
<!-- comment -->
The leaves on the trees in these forests perform photosynthesis, the chemical conversion of CO2 to oxygen.
<!-- comment -->
During the winter months there are no leaves on the trees, so CO2 accumulates in the atmosphere until it peaks in May of each year.
<!-- comment -->
When the leaves come back each year, they perform photosynthesis throughout Spring and Summer, which causes the atmospheric CO2 concentration to drop until it reaches its yearly minimum in October.

<!-- paragraph -->

We say that this data visualization is "static" or "non-interactive" because the reader can view it but can not change what is displayed.
<!-- comment -->
That is fine for medium sized data sets, in which we can see all the details of the data set.
<!-- comment -->
However, as we discuss in the following section, static data visualization is not sufficient to show all the details in larger data sets.

<!-- paragraph -->

### Large data analysis with interactive data visualization {#large-data}

<!-- paragraph -->

Some data sets are so large that it is not possible or desirable to plot all of the data at once in a static data visualization.
<!-- comment -->
For such "large data" sets, traditional approaches to data analysis include summarizing the data, and then visualizing the summary.
<!-- comment -->
However, the summary can be misleading, because it does not show all the details of the original data.
<!-- comment -->
In such situations, "interactive data visualization" becomes useful.

<!-- paragraph -->

First, let us consider a slightly more complicated form of the Keeling Curve data visualization.

<!-- paragraph -->

```{r, echo=FALSE}
TimeSeries <- ggplot()+
  geom_line(aes(
    year.month.POSIXct, ppm),
    size=1,
    data=co2.details)+
  geom_point(aes(
    year.month.POSIXct, ppm, color=month),
    data=co2.details)+
  xlab("Time of measurement")+
  ylab("Atmospheric carbon dioxide (parts per million)")
TimeSeries
```

<!-- paragraph -->

The plot above shows colored points for each month of the year, rather than for only May and September, the yearly local minima that we wanted to emphasize.
<!-- comment -->
Since it lacks this emphasis, this static plot is not as informative as the previous plot.
<!-- comment -->
This is an example where it is not desirable to plot all of the data at once.
<!-- comment -->
We can solve this problem using the following interactive plot.

<!-- paragraph -->

```{r ch01-viz-Keeling, echo=FALSE}
animint(
  TimeSeries,
  first=list(month=c("May", "October")))
```

<!-- paragraph -->

In the plot above, the default emphasis is May and October, but the user can click the legend to update the emphasis.
<!-- comment -->
This simple example illustrates the main idea of interactive data visualization using animint.
<!-- comment -->
There are many choices that must be made to show details of big data sets.
<!-- comment -->
For example, the choice of which months to emphasize in the plot above.
<!-- comment -->
Rather than fixing such choices in a static plot, the goal of interactive data visualization is to allow the reader to see what the plot looks like when different choices are made.
<!-- comment -->
In the example above, we used an interactive legend which allows the user to select different months and see what the plot looks like after changing the selection.

<!-- paragraph -->

The example above also provides a good example of `clickSelects` and `showSelected`, the two keywords that animint introduces to allow interaction.
<!-- comment -->
Without going into too many details, the plot above uses `clickSelects=month` for the interactive legend, meaning that clicking the legend should change the selected months.
<!-- comment -->
Furthermore, we used `showSelected=month` for the points, meaning that we should only plot the set of points which corresponds to the currently selected months.
<!-- comment -->
In Chapters 3-4, we will explain how to design data visualizations by writing R code using these two new keywords.

<!-- paragraph -->

## Chapter summary and exercises {#exercises}

<!-- paragraph -->

This chapter explained some basic facts about data, and gave definitions of different sizes of data: small, medium, and large.

<!-- paragraph -->

- Based on the definitions introduced in this chapter, what is the difference between small and medium data?
<!-- comment -->
- What is the difference between medium and large data?

<!-- paragraph -->

Next, [Chapter 2](ch02) explains how the grammar of graphics can be used to create data visualizations.

<!-- paragraph -->


